---
title: ""
author: "C. A. Hamm"
date: "`r format(Sys.Date())`"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echot = TRUE, fig.align = "center")
```

**Wing shape evolution in *Hamadryas* butterflies corresponds to vertical microhabitat use and species range size**

This document is an `RMarkdown` version of the code for the Hamm, Penz and DeVries manuscript:

```{r load package}
library("geomorph")
library("phytools") 
library("geiger")
library("spaceMovie")
```

# Prepare data 

## Import *Hamadryas* phylogeny

```{r read_tree}
Hama <- read.nexus("../Data/Hamydryas_ml.tre")
plot(Hama)
is.ultrametric(Hama)
```

## Make tree ultrametric

```{r lambda, include = TRUE, results = "hide"}
lam <- 10^(-1:6) # lamdba is the smoothing parameter. 
cv <- sapply(lam, function(x) sum(attr(chronopl(Hama, lambda = x, CV = TRUE), "D2")))
plot(lam, cv, pch = 19, ylab = "cross-validation score", 
	xlab = expression(paste(lambda)), las = 1, cex = 1.5) # lowest CV is 0.1

Hama2 <- chronopl(phy = Hama, lambda = 0.1, CV = TRUE, eval.max = 1e3, iter.max = 1e4)
is.ultrametric(Hama2) # Tree is now ultrametric
```

## Import *forewing* data

```{r import_forewing, include = TRUE, results = "hide"}
fore <- readland.tps("../Data/Hama_dorsal_3_aligned.tps",	specID = "ID")
str(fore)
dim(fore) # 50 landmarks with X & Y coordinates, for 71 samples

fore_gpa <- gpagen(A = fore, PrinAxes = FALSE, Proj = TRUE, ProcD = TRUE, max.iter = 1e4, print.progress = FALSE) # used Procrustes distance for sliding in TPSrelW
plotOutliers(fore_gpa$coords) # laodamia and velutina are a bit away from everyone else, but they are weird to begin with.

fw <- two.d.array(fore_gpa$coords)
dim(fw)
table(rownames(fw))
match(Hama2$tip.label, row.names(fw)) # confirm names in tree and data match
```

## Import *hind wing* data

```{r import_hind, include = TRUE, results = "hide"}
hind <- readland.tps("../Data/Hama_vent_3.tps", specID = "ID")
str(hind)
dim(hind)

hind_gpa <- gpagen(A = hind, PrinAxes = FALSE, Proj = TRUE, ProcD = TRUE, max.iter = 1e4, print.progress = FALSE) # used Procrustes distance for sliding in TPSrelW

plotOutliers(hind_gpa$coords) # laodamia and velutina ARE weird (found exclusively in the canopy), chloe is the earliest diverging taxon in our phylogeny and has weird looking hind wings

hw <- two.d.array(hind_gpa$coords)
dim(hw)
table(rownames(hw))
match(Hama2$tip.label, row.names(hw)) # confirm names in tree and data match

groupH <- as.factor(unique(rownames(hw)))
```

# *Forewing* analysis 

## Visualize data
With the data read in, lets take a look at them
```{r fw_visualize, include = TRUE, results = "hide"}
# examine correlations of characters within wings as well
plotAllSpecimens(fore_gpa$coords, mean = TRUE)
# str(fore_gpa)

plotTangentSpace(fore_gpa$coords, label = TRUE, warpgrids = TRUE) # PC1 40.8%, PC2 25.7%
```

## Mean shape
```{r fw_mean_shape}
# Now calculate mean shape because we can only use one indidivual per species.
groupF <- as.factor(unique(rownames(fw)))

pF <- dim(fore_gpa$coords)[1]
kF <- dim(fore_gpa$coords)[2]
YF <- array(NA, dim = c(pF, kF, length(levels(groupF))))
dimnames(YF)[[3]] <- levels(groupF)
dim(YF)

# A function to calculate mean shape
for(i in 1:length(levels(groupF))){
	grpF <- fw[which(groupF == levels(groupF)[i]), ]
	fooF <- arrayspecs(grpF, pF, kF)
	YF[, , i] <- mshape(fooF)
}

#YF # object with mean shape by species
dim(YF)
plot(YF[, 1, ], YF[, 2, ], pch = 19)

# Convert to 2d array
YF2d <- two.d.array(YF)
#head(YF2d)
dim(YF2d)

# I think that my function worked correctly, but I want to be super duper extra safe, so I'm going to double check. 
velutina_fw <- fw[c(17, 68:70), ] # 4 velutina samples
v1F <- arrayspecs(velutina_fw, p = 50, k = 2)
v2F <- mshape(v1F)

plot(YF[, , "velutina"], pch = 19, las = 1, ylab = "", xlab = "")
points(v2F[, 1], v2F[, 2], pch = 1, col = SW_palette("Inquisitor")[1]) # good, the points match up!

# Now a plot of mean shape by species
plotTangentSpace(A = YF, label = TRUE) # PC1 = 37%, PC2 = 28%
```

## Phylomorphospace
```{r fw_phylo}
plotGMPhyloMorphoSpace(phy = Hama2, A = YF, ancStates = FALSE) 
```

Here we set up the forewing data into three partitions:

1. leading edge
1. outer margin
1. trailing edge

## Sanity check
```{r fw_partitions}
# Now confirm the set of landmarks that correspond to leading, outer, and trailing edge of the wing. The circles should perfectly overlap if I am properly delimiting the data.
plot(YF[, 1, ], YF[, 2, ], pch = 19)
points(YF[1:6, 1, ], YF[1:6, 2, ], col = SW_palette("Zeb")[3], pch = 1) # outer margin of wing
points(YF[7:33, 1, ], YF[7:33, 2, ], pch = 1, col = SW_palette("Zeb")[4]) # leading edge
points(YF[34:50, 1, ], YF[34:50, 2, ], pch = 1, col = SW_palette("Zeb")[1]) # trailing edge points
```

## Centroid size
```{r fw_centroids}
## Import forewing centroid size data
fore_cs <- read.delim("../Data/Hama_dorsal_3_centroid.txt", sep = "\t", header = TRUE)
head(fore_cs)
str(fore_cs)
#fore_cs
table(fore_cs$species)

## calculate mean centrois size by species
fw_cs <- with(fore_cs, tapply(csize, species, mean))
fw_cs <- fw_cs[-4] #remove arete, which is not in the tree
fw_cs <- fw_cs[Hama2$tip.label] #make sure the labels are in the same order as the tree
fw_cs 
```

# *Hind wing* analyses 

## Visualize data
```{r hw_visualize}
plotAllSpecimens(hind_gpa$coords, mean = TRUE)

plotTangentSpace(hind_gpa$coords, label = TRUE, warpgrids = TRUE)
```

## Mean shape
```{r hw_mean_shape}
pH <- dim(hind_gpa$coords)[1]
kH <- dim(hind_gpa$coords)[2]
YH <- array(NA, dim = c(pH, kH, length(levels(groupH))))
dimnames(YH)[[3]] <- levels(groupH)
dim(YH)

for(i in 1:length(levels(groupH))){
	grpH <- hw[which(groupH == levels(groupH)[i]), ]
	fooH <- arrayspecs(grpH, pH, kH)
	YH[, , i] <- mshape(fooH)
}
#YH
dim(YH)
plot(YH[, 1, ], YH[, 2, ], pch = 19)

YH2d <- two.d.array(YH)
#head(YH2d)
dim(YH2d)

# Let's make sure that the above function calulated a mean that looks like the mean calculated for one species by `geomorph`.
velutina_hw <- hw[c(17, 69:71), ] # The 4 velutina samples
v1H <- arrayspecs(velutina_hw, p = 12, k = 2)
v2H <- mshape(v1H)

plot(YH[, , "velutina"], pch = 19, las = 1, ylab = "", xlab = "")
points(v2H[, 1], v2H[, 2], pch = 1, add = TRUE, col = SW_palette("Inquisitor")[1]) # Points match up
```

## Phylomorphospace
```{r Hw_phylo}
plotGMPhyloMorphoSpace(phy = Hama2, A = YH, ancStates = FALSE) 
```

## Sanity check
```{r HW_sanity}
# Now confirm the set of landmarks that correspond to the structural and outer margins of the wings. The circles should perfectly overlap if I am properly delimiting the data. 
plot(YH[, 1, ], YH[, 2, ], pch = 19, ylab = "", xlab = "", las = 1)
points(YH[1:6, 1, ], YH[1:6, 2, ], col = SW_palette("Chopper")[1], pch = 1, add = TRUE) # structural landmarks
points(YH[7:12, 1, ], YH[7:12, 2, ], ylab = "", xlab = "", col = SW_palette("AT_DP")[3], pch = 1, add = TRUE) # wing maring landmarks
```

## Centroid size
```{r hw_centroids}
#hind_gpa$Csize
hind_cs <- as.data.frame(names(hind_gpa$Csize))

hind_cs$Csize <- (hind_gpa$Csize)
hind_cs <- reshape::rename(hind_cs, c("names(hind_gpa$Csize)" = "species"))
#hind_cs
table(hind_cs$species)

hw_cs <- with(hind_cs, tapply(Csize, species, mean))
hw_cs

hw_cs <- hw_cs[Hama2$tip.label] #make sure the labels are in the same order as the tree
hw_cs 
```

# Comparative analyses 

## Covariate data
```{r covariate_data}
# import range data
ranges <- read.csv("../Data/Hamadryas_range.csv", header = TRUE, row.names = 1)
ranges$Range <- as.factor(ranges$Range)
str(ranges)
hist(ranges$Pixels, col = "grey", breaks = "Scott", las = 1, main = "", xlab = "# of pixels") # three breaks with the Scott method: Small <= 200000; 200001 < Medium >= 400000; 400001 >= Large 

H.range <- matrix(ranges$Range, dimnames = list(row.names(ranges)))
H.range <- as.factor(H.range)
names(H.range) <- row.names(ranges)
H.range <- H.range[Hama2$tip.label]
H.range

# Import habitat data
cu <- c(NA, 0, 0.5, 0.5, 0, 0.5, 0, 0, 0, 0.5, 0.5, 0, 0, 0 , NA, 1, 1) #0 = understory, 0.5 = mixed, 1 = canopy
cu <- matrix(cu, dimnames = list(rownames(YF2d)))
cu <-cu[Hama2$tip.label, ]
cu1 <- cu[-c(3, 9)]
cu1

# Prune the tree
Hama.pruned <- treedata(phy = Hama2, data = cu1, sort = TRUE, warnings = TRUE) # Safety check to ensure that the covariate data and tree tips match up
plot(Hama.pruned$phy)
```

## Comparative methods
```{r comparative_analysis}
## Phylogenetic signal for forewing shape
FW_sig <- physignal(phy = Hama2, A = YF, iter = 1e4, seed = 876234872, print.progress = FALSE) # K = 0.716, P << 0.001, FW shape is associated with phylogeny

# phylogenetic signal for forewing centroid size
fw_cs_sig <- physignal(phy = Hama2, A = as.matrix(fw_cs), iter = 1e4, seed = 731427, print.progress = FALSE) # K = 0.30, P = 0.856, centroid size no significant phylogenetic signal

# Phylogenetic signal for hind wing shape
hw_sig <- physignal(phy = Hama2, A = YH, iter = 1e4, seed = 826342, print.progress = FALSE) # K = 0.8, P = 0.0018, significant signal for HW shape

# phylogenetic signal for hid wing centroid size
hw_cs_sig <- physignal(phy = Hama2, A = as.matrix(hw_cs), iter = 1e4, seed = 6445, print.progress = FALSE) # K = 0.314, P = 0.81, no significant signal for HW centroid size
```

## Ancestral state reconstruction
```{r ASR}
# ancestral state reconstruction
anc.ML(tree = Hama.pruned$phy, x = cu1, maxit = 1e4, model = "BM")

fastAnc(tree = Hama.pruned$phy, x = cu1, vars = TRUE, CI = TRUE)# pretty much same estimates for both ASR methods

contMap(Hama.pruned$phy, x = cu1, res = 1000)

contMap(Hama.pruned$phy, x = cu1, res = 1000, type = "fan", legend = FALSE)
```

## Phylogenetic integration
```{r morph_covar}
# leading and trailing edge of FW
LE_TE_pls <- phylo.integration(A = YF[7:33, , ], A2 = YF[34:50, , ], phy = Hama2, iter = 1e4, seed = 98234, print.progress = FALSE) # r-PLS = 0.867, P = 0.001

# internal and terminal (vein) of HW
HW_pls <- phylo.integration(A = YH[1:6, , ], A2 = YH[7:12, , ],	phy = Hama2, iter = 1e4, seed = 876234, print.progress = FALSE) # r-pls = 0.93, P = <<0.001

#accounting for phylogeny, are FW and HW "integrated"?
FW_HW <- phylo.integration(A = YF, A2 = YH, phy = Hama2, iter = 1e4, seed = 987324, print.progress = FALSE) # r-pls = 0.85, P = 0.0028
```

## Phylogenetic independent constrast
```{r pic}
#fore- and hind wing csize covariation? 
#standard phylogenetic regression
plot(x = hw_cs, y = fw_cs, pch = 19, las = 1, ylab = "Forewing centroid size", xlab = "Hind wing centroid size")
fw_pic <- pic(fw_cs, phy = Hama2, scaled = TRUE)
hw_pic <- pic(hw_cs, phy = Hama2, scaled = TRUE)
m1 <- lm(fw_pic ~ hw_pic)
summary(m1)

plot(x = hw_pic, y = fw_pic, pch = 19, las = 1, ylab = "Forewing centroid size", xlab = "Hind wing centroid size")
abline(m1, lwd = 3, lty = 3)

#phylogenetic regression through the origin
lm01 <- lmorigin(fw_pic ~ hw_pic, nperm = 1e4)
lm01
```

## Rates of morphological change

```{r forewing_rates}
# forewing, change rate based on habitat
fw_habitat_rate <- compare.evol.rates(phy = Hama2, A = YF, gp = cu1, iter = 1e4, print.progress = FALSE) 
fw_habitat_rate # no association with habitat
fw_habitat_rate$pairwise.pvalue

#forewing, change based on centroid size
fwcs_habitat_rate <- compare.evol.rates(phy = Hama2, A = as.matrix(fw_cs), gp = cu1, iter = 1e4, print.progress = FALSE)
fwcs_habitat_rate # no association with centroid size

# forewing, rate based on range size
fw_range_rate <- compare.evol.rates(phy = Hama2, A = YF, gp = H.range, iter = 1e4, print.progress = FALSE) 
fw_range_rate # no sig difference in FW rates based on habitat size

# forewing centroid size and range size
fwcs_range_rate <- compare.evol.rates(phy = Hama2, A = as.matrix(fw_cs), gp = H.range, iter = 1e4, print.progress = FALSE)
fwcs_range_rate # no association
```

## *Hind wing* change
```{r hind_wing_rates}
# rate of HW change based on habitat
Hw_habitat_rate <- compare.evol.rates(phy = Hama2, A = YH, gp = cu1, iter = 1e4, print.progress = FALSE)
Hw_habitat_rate # Yup. HW has different rates based on habitat
Hw_habitat_rate$pairwise.pvalue # can't distinguish 0-0.5, can distinguish canopy from both understory and mixed. That is cool, considering FW and HW are integrated, but HW rates faster in canopy. 

# hind wing centroid size based on habitat
Hwcs_habitat_rate <- compare.evol.rates(phy = Hama2, A = as.matrix(hw_cs), gp = cu1, iter = 1e4, print.progress = FALSE)
Hwcs_habitat_rate # No sig association

# rate of HW change based on range
Hw_range_rate <- compare.evol.rates(phy = Hama2, A = YH, gp = H.range, iter = 1e4, print.progress = FALSE) 
Hw_range_rate # HW rate is sig fast
Hw_range_rate$pairwise.pvalue

Hwcs_range_rate <- compare.evol.rates(phy = Hama2, A = as.matrix(hw_cs), gp = H.range, iter = 1e4, print.progress = FALSE) # large and medium ranges are faster than small, large MUCH faster
```

##  Habitat plot
```{r plots}
nspecies <- length(Hama2$tip.label)
cu_color <- cu
cu_color[cu == 0.0] <- SW_palette("TFA")[7]
cu_color[cu == 0.5] <- SW_palette("TPM")[8]
cu_color[cu == 1.0] <- SW_palette("ROTJ")[5]
cu_color[is.na(cu)] <- SW_palette("Inquisitor")[3]

plot.phylo(Hama2, font = 3, label.offset = 0.07, edge.width = 3)
points(x = rep(1.035, nspecies), y = 1:nspecies, pch = 19, col = cu_color, cex = 1.5)
legend("topleft", legend = c("Canopy", "Mixed", "Ground", "Unknown"), pch = 19, col = c(SW_palette("ROTJ")[5], SW_palette("TPM")[8], SW_palette("TFA")[7], SW_palette("Inquisitor")[3]), pt.cex = 1.5, bty = "n", title = expression(bold("Habitat")))
```
